---
title: "Bayesian Deep Learning and Approximate Bayesian Computation for parameter inference"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{parameter_inference}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo = F, eval = F}
# Install ABCNeuralNet
devtools::install_github("ThomasBrazier/abcneuralnet")

# Install C++ dependencies of torch (e.g. CUDA, lantern)
torch::install_torch()
# After installing torch CUDA dependencies, the R session must be reloaded
```


```{r setup3, echo = F}
library(abcneuralnet)
library(ggplot2)
library(torch)
library(tidyverse)
library(kableExtra)
```


# Introduction


Approximate Bayesian Computation (ABC) has emerged as a powerful framework for parameter inference in complex statistical models where likelihood functions are intractable. However, traditional ABC methods face significant computational challenges when dealing with high-dimensional summary statistics and complex parameter spaces. This vignette presents **abcneuralnet**, an R package that integrates Bayesian deep learning with ABC to address these challenges. We demonstrate four methodological approaches: Concrete Dropout [@gal2017concrete], Monte Carlo Dropout [@gal2016dropout], Deep Ensemble [@lakshminarayanan_simple_2016], and TabNet-ABC [@arik_tabnet_2020; @akesson_convolutional_2022]. Each method provides different strategies for uncertainty quantification and parameter inference, combining the representational power of deep neural networks with principled Bayesian inference. Through comprehensive examples, we illustrate the theoretical foundations, practical implementation, and performance characteristics of these methods for parameter inference in population genetics and related fields.


## The challenge of parameter inference in complex population genetics models

Modern population genomics inference increasingly relies on complex stochastic models for which likelihood functions are either analytically intractable or computationally prohibitive to evaluate. This challenge is particularly acute in population genetics, evolutionary biology, and systems biology, where both coalescent and forward-time simulators can generate data under sophisticated models, but the corresponding likelihood functions remain intractable [@beaumont_approximate_2019].

Approximate Bayesian Computation (ABC) provides a likelihood-free framework for parameter inference in such settings. The fundamental principle of ABC relies on the intuition that simulated data sets with summary statistics close to those observed are likely to have been generated by similar parameter values. Formally, for observed data $\mathcal{D}_\text{obs}$ with summary statistics $S(\mathcal{D}_\text{obs}) = s_\text{obs}$, ABC rejection sampling accepts parameters $\theta$ from prior $\pi(\theta)$ if the distance $\rho(S(\mathcal{D}(\theta)), s_\text{obs}) < \epsilon$ for some tolerance threshold $\epsilon$.


## Limitations of traditional ABC approaches

Despite its theoretical simplicity and elegance, traditional ABC faces several practical challenges:

1. **Curse of dimensionality**: The acceptance rate decreases exponentially with the number of summary statistics [@blum2010approximate]
2. **Distance metric selection**: The choice of the distance function $\rho(\cdot,\cdot)$ significantly impacts performance [@joyce2008approximately]
3. **Tolerance specification**: Setting appropriate $\epsilon$ values involves trade-offs between accuracy and computational efficiency
4. **Computational efficiency**: Rejection sampling can be highly inefficient, especially for complex posteriors. In addition, traditional ABC do no scale well for multiple observed samples.


## Deep Learning as a solution

Recent advances in deep learning offer promising solutions to these challenges [@baragatti_approximate_2024]. Neural networks can learn complex mappings between high-dimensional summary statistics and parameter spaces, effectively learning an implicit likelihood or posterior approximation.

The **abcneuralnet** package implements four state-of-the-art Bayesian deep learning approaches:

1. **Concrete Dropout**: Provides rigorous uncertainty quantification through approximate variational inference, specifically a refinement of “dropout as variational Bayesian approximation” where the dropout rate is learned during inference [@gal2017concrete; @baragatti_approximate_2024]
2. **Monte Carlo Dropout**: A simpler "dropout" approach, which offers uncertainty estimation through stochastic forward passes, yet requiring careful tuning of the dropout rate hyper-parameter [@gal2016dropout]
3. **Deep Ensemble**: A highly efficient ensemble approach delivering robust predictions through model averaging [@lakshminarayanan_simple_2016]
4. **TabNet-ABC**: Combines interpretable attention mechanisms with traditional ABC regression [@arik_tabnet_2020; @akesson_convolutional_2022]

Each method addresses different aspects of the ABC inference problem, with their own advantages and limitations, while maintaining compatibility with existing ABC workflows and providing rigorous uncertainty quantification essential for scientific inference.



Throughout this tutorial, I will use the three core methods developed in **abcneuralnet**, specifically **Concrete Dropout**, **Deep Ensemble** and **Tabnet-ABC**, to show how to perform end-to-end parameter inference on simple toy models: initialize an `abcnn` object, fit the neural network on simulated data, predict on test pseudo-observed, predict parameters on observed samples, and finally evaluate the performance of the trained neural network on both test and observed data.


# Computational Implementation

## Architecture and Dependencies

The **abcneuralnet** package is built on the **torch** ecosystem for R [@torch2025], providing efficient tensor operations and automatic differentiation. The key architectural components include:

- **R6 abcnn()**: the core R6 class implementing all methods and functions in a single R object: fit, predict and plot results
- **Core Engine**: `torch::nn_module` implementing different Bayesian neural network methods
- **Training Framework**: Integration with **luz** for high-level training loop management []
- **Explainability and Interpretability methods**: cutting-edge explainability methods in `explainn()` with the package **innsight** [@koenen_interpreting_2024]
- **Calibrated Conformal Prediction**: Calibrated Credible Intervals are obtained with Conformal Prediction as described in [@baragatti_approximate_2024].
- **Model Persistence**: Comprehensive serialization through **bundle** and native torch save mechanisms with `save_abcnn()` and `load_abcnn()`


If you are a beginner with `torch` in R and want to know more about basic and more advanced usages, I recommend you to read the book ['Deep Learning and Scientific Computing with R torch'](https://skeydan.github.io/Deep-Learning-and-Scientific-Computing-with-R-torch/) by Sigrid Keydana.


## Hardware

### GPU Acceleration

The package automatically detects and utilizes CUDA-enabled GPUs through the **luz** backend. GPU acceleration provides speedups of 10-100x for training, particularly beneficial for:

- Large-scale ensemble training (5-10 networks) and deeper networks (high number of layers and neurons)
- High-dimensional summary statistics (>100 dimensions) and large-scale datasets (>$10^6$ training simulations)
- Extensive hyperparameter tuning


**abcneuralnet** uses the `torch` R package, which is a very efficient implementation leveraging the Torch C++ library and providing codes and functions very similar to the `PyTorch` framework. Neural networks are fitted with the `luz` R package. If you are beginners with `torch` in R and want to know more about it, I recommend you to read the book ['Deep Learning and Scientific Computing with R torch'](https://skeydan.github.io/Deep-Learning-and-Scientific-Computing-with-R-torch/) by Sigrid Keydana.


# Data analysis workflow

**abcneuralent** is designed to be user-friendly and streamline the data analysis workflow:


(1) Initialize and `abcnn` `R6` object with the `abcnn$new()` method, providing the training simulated dataset (the reference table, with simulated parameters and summary statistics). You also need to set the hyperparameters of the neural network at this step (e.g. number of layers, neurons, epochs, the learning rate), though you can change them at any time by accessing `R6` public slots.

(2) Fit the neural net with `abcnn$fit()`. The **luz** engine is used to fit the model.

(3) Assess the quality of the fit, by visually assessing the training curve with `abcnn$plot_training()`.

(4) Estimate the cross-validation error with `abcnn$cross_validation(cross_validation_param, cross_validation_sumstats)`, where `cross_validation_param` and `cross_validation_sumstats` are an unseen pseudo-observed dataset (i.e simulations) for which ground truth is known. The function returns a table with performance metrics. Once cross-validation has been computed, you can call it back without arguments (i.e. `abcnn$cross_validation()`) to get the metrics table, or plot the cross-validation scatter plot with `abcnn$plot_cross_validation()`. Alternatively, predict on test samples with `abcnn$predict()` to evaluate the posterior predictive performance with your custom method. If necessary, the test samples can be changed at this step with `abcnn$predict(data = new_data)`, so that you can predict for different dataset with the same trained neural network.

(5) Assess the goodness-of-fit on test samples with the graphical outputs of `abcnn$plot_cross_validation()`, `abcnn$plot_prediction()` and `abcnn$plot_posterior()`.

(6) Predict on observed samples with `abcnn$predict(data = observed_data)` and look at estimates with `abcnn$plot_posterior()` (figure) and `abcnn$predictions()` (table).

(7) Perform posterior predictive check. Draw parameters for `n` random samples in the posterior distribution with `abcnn$draw_from_posterior(n)`. Re-simulate summary statistics under your generative model with the parameters drawn from the posterior. Compare the fit of the distribution of your observed summary statistics with the distribution of re-simulated summary statistics given the posterior distributions.



# Case study 1: Linear Regression with uncertainty quantification


## Problem setup

We begin with a simple linear regression problem to demonstrate fundamental concepts:

$$y = wx + b + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2)$$

where $w = 2$, $b = 8$, and $\sigma = 1$. This controlled setting allows us to validate uncertainty quantification against analytical solutions.


## Data generation

I simulated 2,000 training samples and 1,000 test samples with a simple linear function. There is only one parameter to estimate, and one input variable.


```{r toy_data_1, echo = F, eval=T}
n_train = 100000 # Number of data points
n_obs = 1000 # Validation size

gen_data_1d = function(n) {
  sigma = 1
  X = matrix(rnorm(n))
  w = 2
  b = 8
  Y = matrix(X %*% w + b + sigma * rnorm(n))
  list(X = X, Y = Y)
}

XY = gen_data_1d(n_train + n_obs)

X_train = XY$X[1:n_train]
Y_train = XY$Y[1:n_train]

X_obs = XY$X[(n_train + 1):(n_train + n_obs)]
Y_obs = XY$Y[(n_train + 1):(n_train + n_obs)]

# Predict Y when X is observed
theta = data.frame(y1 = Y_train)
sumstats = data.frame(x1 = X_train)
observed = data.frame(x1 = X_obs)

df_concrete = list(X_train = X_train,
                   Y_train = Y_train,
                   X_obs = X_obs,
                   Y_obs = Y_obs)
```


```{r save_toy1, echo=F, eval = F}
# Save the dataset
saveRDS(df_concrete, "../inst/extdata/df_concrete.rds")
```


```{r load_toy1, echo=F}
# Load it back
df_concrete = readRDS("../inst/extdata/df_concrete.rds")
```


```{r dataset_toy1, echo=F}
X_train = df_concrete$X_train
Y_train = df_concrete$Y_train

X_obs = df_concrete$X_obs
Y_obs = df_concrete$Y_obs

theta = data.frame(y1 = Y_train)
sumstats = data.frame(x1 = X_train)
observed = data.frame(x1 = X_obs)
```


## Fit a model with **Concrete Dropout**

I used a Concrete Dropout regression model described in [@gal2017concrete] and applied conformal prediction as in [@baragatti_approximate_2024]. You can read an interesting discussion on the Concrete Dropout model on this [blog post](https://blogs.rstudio.com/ai/posts/2018-11-12-uncertainty_estimates_dropout/) for more details.


I initialized the `abcnn` object with the training samples `theta` (parameters of the simulations), `sumstats` (the summary statistics) and `observed` (the summary statistics of the test samples, also simulated). For this simple toy model, I do not have true observed data to predict at the end. I implemented a three-layer network with 256 hidden units per layer:


```{r, echo = T}
# Init an abcnn object with inputs and targets
abc = abcnn$new(theta,
            sumstats,
            observed,
            method = 'concrete dropout',
            scale_input = "none",
            scale_target = "none",
            num_hidden_layers = 3,
            num_hidden_dim = 128,
            epochs = 20,
            batch_size = 32,
            l2_weight_decay = 1e-5,
            learning_rate = 0.001,
            seed = 6295)
```


You can have a summary of your `abcnn` object at any time with `abcnn$summary()`.


```{r, echo = T}
abc$summary()
```


## Model training and performance



```{r, echo = T, eval = F}
# Use the fit() method to train the neural network
abc$fit()
```


```{r, echo=F, eval= T}
# save_abcnn(abc, prefix = "../inst/extdata/abc_concrete")

abc = load_abcnn(prefix = "../inst/extdata/abc_concrete")
```


The `abcnn` object is a `R6Class` object, providing a modern Object-Oriented Programming (OOP) approach. All the settings and data of an `abcnn` object can be accessed in public slots. For example, the `torch` model can be accessed with `abcnn$fitted$model`, and the `luz` fitted model can be accessed with `abcnn$fitted`. It means that `abcnn` objects allows very flexible approaches, where you can plot and visualize data and outputs, and define new functions easily. For example, I could wish to plot the evolution of the loss value at each iteration (epoch) during the training. This is the regular training curve used to assess training quality in every deep learning model.

The `abcnn` object is an R6 class that uses a modern object‑oriented design in R. All its settings and data are stored in public fields that you can access directly. For instance, the underlying torch model is in `abcnn$fitted$model`, and the **luz** fitted model is in `abcnn$fitted`. This structure makes `abcnn` objects very flexible, so you can easily visualize data and outputs or write new helper functions, such as plotting the loss value at each training epoch to inspect the training curve of the deep learning model.


```{r, echo = T}
# The torch model
# abc$fitted$model

# The luz fitted model
abc$fitted

# The number of dimensions (i.e. neurons) and layers
abc$num_hidden_dim
abc$num_hidden_layers
```





Plotting the training curve is a really common procedure for fitting neural network. It gives you a lot of information on the quality of the training, especially to assess over- and under-fitting of the network. Read more about the [bias-variance tradeoff](). Training a neural network usually involves a trial-error process with iterative hyperparameter tuning until a good performance is achieved, with a good balance between over- and under-fitting.


Plot the training curve of the deep learning model to check model fit:


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center", fig.cap="The training curve of the neural network across 30 epochs, computed on training data split in three partitions: training, validation (also called testing) and evaluation. Training and validation are computed at the end of each epoch. The black horizontal line is the loss computed on the evaluation dataset at the end of the training procedure."}
abc$plot_training()
```


Note that sometimes the heteroscedastic loss value can be negative with the Concrete Dropout method. It can be confusing as MSE losses are always positive. However, negative concrete dropout losses do not impact the training procedure.

You can save the `abccn` object with the fitted model and predictions.


```{r, echo = T, eval = F}
save_abcnn(abc, prefix = "../path/abc_concrete")

abc = load_abcnn(prefix = "../path/abc_concrete")
```


## Cross-validation


Re-use the generative model to generate a new dataset of unseen simulations with ground truth:


```{r toy_data_2, echo = F, eval = F}
n_crossval = 1000 # Validation size

crossval = gen_data_1d(n_crossval)

theta_crossval = data.frame(y1 = crossval$Y)
sumstats_crossval = data.frame(x1 = crossval$X)
```


Compute the cross-validation error metrics:

```{r, echo = T, eval = F}
abc$cross_validation(theta_crossval,
                     sumstats_crossval)
```


Print the already computed cross-validation error metrics:

```{r, echo = T}
abc$cross_validation()
```


Plot the cross-validation scatter plot (predicted ~ ground truth) with Conformal Credible Intervals:

```{r, echo = T}
abc$plot_cross_validation()
```


## Estimates and uncertainty with the **Concrete Dropout** method


After the model is efficiently trained and the training quality has been assessed, parameter estimates can be predicted with `abc$predict()`. I predicted parameters for 1,000 simulated samples (summary statistics):


```{r, echo = T, eval = F}
abc$predict()
```


```{r, echo=F, eval=F}
save_abcnn(abc, prefix = "../inst/extdata/abc_concrete")
```




You can get predictions with their associated uncertainties and Conformal Credibility Intervals as a tidy data frame:

```{r, echo = T, message=F}
head(abc$predictions())
```


The trained model provides separate estimates of aleatoric and epistemic uncertainty:


```{r, echo = F}
df_predicted = abc$predictions()

df_predicted$uncertainty_a_upper = df_predicted$predictive_mean + df_predicted$aleatoric_uncertainty
df_predicted$uncertainty_a_lower = df_predicted$predictive_mean - df_predicted$aleatoric_uncertainty

df_predicted$uncertainty_e_upper = df_predicted$predictive_mean + df_predicted$epistemic_uncertainty
df_predicted$uncertainty_e_lower = df_predicted$predictive_mean - df_predicted$epistemic_uncertainty

df_predicted$ci_conformal_upper = df_predicted$predictive_mean + df_predicted$overall_conformal_credible_interval
df_predicted$ci_conformal_lower = df_predicted$predictive_mean - df_predicted$overall_conformal_credible_interval

df_predicted$ci_conformal_e_upper = df_predicted$predictive_mean + df_predicted$epistemic_conformal_credible_interval
df_predicted$ci_conformal_e_lower = df_predicted$predictive_mean - df_predicted$epistemic_conformal_credible_interval

df_predicted$x = X_obs
df_predicted$y_true = Y_obs

df_training = data.frame(x = X_train,
                         y = Y_train)
```

```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center", fig.cap="Predictions as a function of simulated parameter. The purple ribbon is the Conformal Credible Interval based on the epistemic unvertainty alone. The green ribbon is the Conformal Credible Interval based on the overall unvertainty."}
ggplot(data = df_training[1:1000,], aes(x = x, y = y)) +
  geom_point(color = "blue", alpha = 0.3) +
  # geom_point(data = df_predicted, aes(x = x, y = y_true), color = "green", alpha = 0.3) +
  geom_line(data = df_predicted, aes(x = x, y = predictive_mean), color = "Red") +
  geom_point(data = df_predicted, aes(x = x, y = predictive_mean), color = "Red") +
  facet_wrap(~ parameter, scales = "free") +
  geom_ribbon(data = df_predicted, aes(x = x, y = predictive_mean, ymin = ci_conformal_e_upper, ymax = ci_conformal_e_lower), alpha = 0.4, fill = "purple") +
  geom_ribbon(data = df_predicted, aes(x = x, y = predictive_mean, ymin = ci_conformal_upper, ymax = ci_conformal_lower), alpha = 0.3, fill = "green") +
  theme_bw()
```

There are different Credible Intervals and uncertainty measures provided. I recommend to use the overall Conformal Credible Interval in practice, which is estimated with Conformal Prediction and using the overall (aleatoric + epistemic) uncertainty as a proxy of the variance (see Baragatti et al 2024 for details on the method). The overall Conformal Credible Interval has a smoothing effect and fits really well the expected C.I. in benchmarks (Brazier, unpublished). If the neural network learned well from the data, both epistemic and overall Conformal Credible Intervals should converged. Yet the epistemic Conformal Credible Interval has an interesting property. It increases more than the overall C.I. in regions where the data has not been seen by the network during the training or where the predictive power is lower than the rest of the dataset, yielding more conservative C.I. in these regions, and giving a useful information for interpreting the model's outputs (see Figure above, especially compare the purple and green C.I.). To validate the fit of the network, it is interesting to look directly to the epistemic and aleatoric uncertainties estimated by Concrete Dropout (see Figure below). Concrete Dropout and Deep Ensemble jointly estimate the mean and the variance of each data point. The uncertainty is the square root of the variance estimated (i.e., standard deviation).



```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center", fig.cap="Predictions as a function of simulated parameters. The epistemic uncertainty (red) and aleatoric uncertainty (blue) were estimated with Concrete Dropout (Gal et 2017)."}
ggplot(data = df_training[1:1000,], aes(x = x, y = y)) +
  geom_point(color = "grey", alpha = 0.3) +
  # geom_point(data = df_predicted, aes(x = x, y = y_true), color = "green", alpha = 0.3) +
  geom_line(data = df_predicted, aes(x = x, y = predictive_mean), color = "Red") +
  geom_point(data = df_predicted, aes(x = x, y = predictive_mean), color = "Red") +
  facet_wrap(~ parameter, scales = "free") +
  geom_ribbon(data = df_predicted, aes(x = x, y = predictive_mean, ymin = uncertainty_a_lower, ymax = uncertainty_a_upper), alpha = 0.3, fill = "blue") +
  geom_ribbon(data = df_predicted, aes(x = x, y = predictive_mean, ymin = uncertainty_e_lower, ymax = uncertainty_e_upper), alpha = 0.3, fill = "red") +
  theme_bw()
```

The aleatoric uncertainty is constant across the dataset, as the variance in the data is homogeneous (i.e. homoscedasticity), while the epistemic uncertainty, reflecting model's uncertainty, decreases at the iddle of the distribution, where the model's predictive performance is better.



The raw uncertainty estimated with **Concrete Dropout**:

```{r concrete_prediction, echo = T, fig.height = 4, fig.width = 8, fig.cap="Prediction results with uncertainty quantification. Red line shows mean predictions, blue ribbon indicates aleatoric uncertainty, and red ribbon shows epistemic uncertainty."}
abc$plot_prediction(uncertainty_type = "uncertainty")
```

The epistemic uncertainty increases at the edges of the distribution of the training dataset.


The Conformal Credible intervals provide calibrated uncertainty estimates with guaranteed coverage properties [@baragatti_approximate_2024]:

```{r concrete_conformal, echo = T, fig.height = 4, fig.width = 8, fig.cap="Conformal credible intervals (green) compared to epistemic uncertainty intervals (purple). Conformal intervals provide calibrated coverage guarantees."}
abc$plot_prediction(uncertainty_type = "conformal")
```

The epistemic Conformal Credible Intervals, based on the epistemic uncertainty as a proxy of uncertainty, are very sensitive to variance in uncertainty, and increases shaprly at the edges of the data distribution. It is a good choice if you are interested in very conservative C.I. and want to assess the sensitivity to over-fitting.


On the contrary, the overall Conformal Credible Intervals, based on both the aleatoric and epistemic uncertainties a proxy of the whole uncertainty, is smoother and less impacted by local variance in epistemic uncertainty. It is a good choice if you are searching for a reliable and well-calibrated C.I.

Interestingly, the epistemic uncertainty increases in regions where the cross-validation plot shows higher error rates (compare the plots above and below):

```{r, echo = T}
abc$plot_cross_validation()
```




If you have the same number of input and output parameters (e.g., x1 and x2 as input, y1 and y2 as output), then the default representation in `abc$plot_prediction()` will naturally represent linear regressions (as many subplots as output parameters). Otherwise predictions are simply represented side by side and the X axis is meaningless (the `plot_type = "errorbar"` option offers a better representation).


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
abc$plot_prediction(uncertainty_type = "uncertainty", plot_type = "errorbar")
```



## Plotting posteriors for individual observations


For specific observations, we can examine the approximate posterior distribution:


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
abc$plot_posterior(sample = 700, prior = TRUE, uncertainty_type = "uncertainty") +
  geom_vline(xintercept = Y_obs[700], color = "red", size = 1.5)
Y_obs[700]
abc$predictive_mean$y1[700]
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center", fig.cap="Distribution of approximate posterior estimates with predictive means and credible intervals. The prior distribution is plotted underneath (white bars) to compare priors and posteriors."}
abc$plot_posterior(sample = 501, prior = TRUE, uncertainty_type = "conformal") +
  geom_vline(xintercept = Y_obs[501], color = "red", size = 1.5)
Y_obs[501]
abc$predictive_mean$y1[501]
```




# Case study 2: Nonlinear multivariate regression with **Deep Ensemble**


## A complex function combining sinus and cosinus functions

To demonstrate the capabilities of Deep Ensemble methods, we consider a challenging nonlinear function with two output parameters:

$$\begin{align}
y_1 &= \sin(x_1) + \epsilon_1 \\
y_2 &= \cos(x_2) + \epsilon_2
\end{align}$$

where $\epsilon_1, \epsilon_2$ represent heteroscedastic noise with varying magnitude across different regions of the input space.


## Data Generation with Missing Regions

We simulate two parameters and two summary statistics, with missing data (gaps) and different amounts of random noise in the two parts of the dataset:

```{r, echo = T, eval = F}
# Parameters of simulated input x
data_range = 7
data_step = 0.0005

# Boundaries of the gap in the data range
bound1 = -2
bound2 = 2

# Random noise applied on y
data_sigma1a = 0.1
data_sigma2a = 0.5

data_sigma1b = 0.2
data_sigma2b = 0.1

# Number of simulated data points
# num_data = 10000

# Simulate x1
data_x1a = seq(-data_range, bound1 + data_step, by = data_step)
data_x1b = seq(bound2, data_range + data_step, by = data_step)
# Simulate targets y
data_y1a = sin(data_x1a) + rnorm(length(data_x1a), 0, data_sigma1a)
data_y1b = sin(data_x1b) + rnorm(length(data_x1b), 0, data_sigma2a)

# Shift X1 to get X2
data_x2a = data_x1a + 7
data_x2b = data_x1b + 7
# Simulate targets y
data_y2a = cos(data_x2a) + rnorm(length(data_x2a), 0, data_sigma1b)
data_y2b = cos(data_x2b) + rnorm(length(data_x2b), 0, data_sigma2b)

df = data.frame(x1 = c(data_x1a, data_x1b),
                    x2 = c(data_x2a, data_x2b),
                    y1 = c(data_y1a, data_y1b),
                    y2 = c(data_y2a, data_y2b))

# Shuffle data
shuffle_idx = sample(1:(nrow(df)), nrow(df), replace = FALSE)
df_train = df[shuffle_idx,]

# Train/Test datasets
# test_ratio = 0.1
# num_train_data = round(nrow(df) * (1 - test_ratio), digits = 0)
# num_test_data  = nrow(df) - num_train_data
# 
# train_x = df[1:num_train_data, c("x1", "x2")]
# train_y = df[1:num_train_data, c("y1", "y2")]
# test_x = df[num_train_data:nrow(df_train), c("x1", "x2")]
# test_y = df[num_train_data:nrow(df_train), c("y1", "y2")]
train_x = df_train[, c("x1", "x2")]
train_y = df_train[, c("y1", "y2")]


# Make a pseudo-obseerved dataset with out of distribution data points
# Simulate x1
data_x1 = seq(-data_range, data_range, length.out = 1000)
# Simulate true targets y
data_y1 = sin(data_x1)

# Shift X1 to get X2
data_x2 = data_x1 + 7
# Simulate targets y
data_y2 = cos(data_x2)

df_observed = data.frame(x1 = data_x1,
                x2 = data_x2,
                y1 = data_y1,
                y2 = data_y2)


observed_x  = df_observed[, c("x1", "x2")]
observed_y  = df_observed[, c("y1", "y2")]

df_deepensemble = list(df_train = df_train,
                       df_observed = df_observed)
```


```{r, echo = F, eval = F}
# Save the dataset
saveRDS(df_deepensemble, "../inst/extdata/df_deepensemble.rds")
```

```{r, echo = F}
# Load it back
df_deepensemble = readRDS("../inst/extdata/df_deepensemble.rds")

df_train = df_deepensemble$df_train
df_observed = df_deepensemble$df_observed

train_x = df_train[, c("x1", "x2")]
train_y = df_train[, c("y1", "y2")]
observed_x  = df_observed[, c("x1", "x2")]
observed_y  = df_observed[, c("y1", "y2")]
```




```{r, echo = F, fig.height = 4, fig.width = 8, fig.align="center"}
# Plot the simulated data
p1 = ggplot(data = df_train, aes(x = x1, y = y1)) +
  geom_point(color = "Blue", alpha = 0.2) +
  geom_line(data = df_observed, aes(x = x1, y = y1), color = "red") +
  theme_bw()

p2 = ggplot(data = df_train, aes(x = x2, y = y2)) +
  geom_point(color = "Green", alpha = 0.2) +
  geom_line(data = df_observed, aes(x = x2, y = y2), color = "red") +
  theme_bw()

ggpubr::ggarrange(p1, p2, ncol = 2)
```




## Deep Ensemble Training

Deep Ensemble trains multiple networks with different random initialization and random batches (adversarial training not implemented yet) [@lakshminarayanan_simple_2016]:


```{r, echo = T}
# devtools::load_all()
# Predict Y when X is observed
theta = train_y
sumstats = train_x
observed = observed_x
true_param = observed_y

abc_ensemble = abcnn$new(theta,
            sumstats,
            observed,
            method = 'deep ensemble',
            num_networks = 5,
            scale_input = "minmax",
            scale_target = "none",
            epochs = 20,
            num_hidden_layers = 3,
            num_hidden_dim = 512,
            batch_size = 128,
            seed = 4722)
```


```{r, echo = T, eval = F}
abc_ensemble$fit()
```


```{r, echo = F}
# save_abcnn(abc_ensemble, prefix = "../inst/extdata/abc_ensemble")

abc_ensemble = load_abcnn(prefix = "../inst/extdata/abc_ensemble")
```


Checking the model training:

```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
abc_ensemble$plot_training()
```





<!-- ```{r, echo = T} -->
<!-- abc_ensemble$cross_validation(true_param[1:100,], -->
<!--                               observed[1:100,]) -->

<!-- abc_ensemble$plot_cross_validation() -->
<!-- ``` -->







## Variation in Uncertainty Quantification across different regions


```{r, echo = T, eval = F}
abc_ensemble$predict()
```

```{r, echo = F, eval = F}
save_abcnn(abc_ensemble, prefix = "../inst/extdata/abc_ensemble")
```



Deep Ensembles provide particularly insightful uncertainty estimates across different data regions:


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
abc_ensemble$plot_prediction(uncertainty_type = "uncertainty")
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
abc_ensemble$plot_prediction(uncertainty_type = "conformal")
```


The uncertainty, especially epistemic uncertainty, increases in unseen regions during training. In addition, the aleatoric uncertainty increases in regions with higher variance (i.e. noise) in data:


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
# Print a sample with -5 < x1 < -4 (within the distribution with a low noise)
# which(abc$observed < -4 & abc$observed > -5)
abc_ensemble$plot_posterior(sample = 155, prior = TRUE)
```

```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
# Print a sample with 4 < x1 < 5 (within the distribution with a high noise)
# which(abc$observed < 5 & abc$observed > 4)
abc_ensemble$plot_posterior(sample = 800, prior = TRUE)
```

```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
# Print a sample with -1 < x1 < 1 (out of training distribution)
# which(abc$observed < 1 & abc$observed > -1)
abc_ensemble$plot_posterior(sample = 520, prior = TRUE)
```







# Case Study 3: A high-dimensional dataset

## Gaussian Toy Model

We demonstrate this approach on a homoscedastic normal model, where $\text{Inv-Gamma}(\alpha, \beta)$ denotes an inverse Gamma distribution with shape parameter $\alpha$ and scale parameter $\beta$:

$$y \sim \mathcal{N}(\mu, \sigma^2), \quad \mu \sim \mathcal{N}(0, \sigma^2), \quad \sigma^2 \sim \text{Inv-Gamma}(\alpha, \beta)$$

It is straightforward to derive the exact values of $E(θ1 | y)$, $Var(θ1 | y)$, $E(θ2 | y)$, $Var(θ2 | y)$ and posterior quantiles for the two parameters from the above expressions and for a given sample y (see [@raynal2019abc] for details on the method).

In addition we add 50 noisy random summary statistics simulated according to a uniform(0,1) distribution to see how the model can handle noise in high dimensional dataset. Hence there are a total of 61 summary statistics, and only 11 of these are informative.



```{r, eval=F, echo=F}
########################################################
#               Simulation of                          #
#           a Gaussian toy example                     #
########################################################

library(mvtnorm) # for multivariate normal distribution
library(spatstat) # for weighted.quantile function
library(MCMCpack) # for the inverse gamma distribution

#====================================
# Homoscedastic Normal toy model
#====================================
# Sample size of y
n = 10
# Inverse gamma parameters
alpha = 4
beta = 3
# Training set sample size
N = 20000
# Test set sample size
p = 1000
# offset for the out-of-dist data
offset = 0

homoscedastic_normal = function(alpha = 4,
                                beta = 3,
                                n = 10,
                                N = 10000,
                                p = 1000,
                                offset = 0) {
  # Function to compute quantiles from
  # student distribution
  qnst = function(p, deg, loca, scale) {
    return(loca + scale * qt(p, df = deg))
  }
  
  # Simulation of the ABC reference table
  
  set.seed(1) # for reproducibility 
  
  # X parameter
  theta1.train = rep(NA, N)
  theta2.train = 1 / rgamma(N, shape = alpha, rate = beta)
  for (i in 1:N) {
    theta1.train[i] = rnorm(1, 0, sqrt(theta2.train[i]))
  }
  
  # Add offset for out of dist
  theta2.train = theta2.train + offset
  theta1.train = theta1.train + offset
  
  # Y parameter
  # n sample cols
  # n training set rows
  y.ref = matrix(NA, N, n)
  for (i in 1:N) {
    y.ref[i, ] = rnorm(n, theta1.train[i], sqrt(theta2.train[i]))
  }
  
  # Compute some summary statistics 
  sumstats.train = matrix(NA, N, 3)
  
  for (i in 1:N) {
    sumstats.train[i, ] = c(mean(y.ref[i, ]), var(y.ref[i, ]), mad(y.ref[i, ]))
  }
  
  ref.training = cbind(theta1.train, theta2.train, sumstats.train)
  colnames(ref.training) = c("theta1", "theta2", "expectation", "variance", "mad")
  
  # Simulation of the ABC test table
  theta1.test = rep(NA, p)
  
  theta2.test = 1 / rgamma(p, shape = alpha, rate = beta)
  for (i in 1:p) {
    theta1.test[i] = rnorm(1, 0, sqrt(theta2.test[i]))
  }
  
  theta2.test = theta2.test + offset
  theta1.test = theta1.test + offset
  
  y.test = matrix(NA, p, n)
  
  for (i in 1:p) {
    y.test[i, ] = rnorm(n, theta1.test[i], sqrt(theta2.test[i]))
  }
  
  # Compute some summary statistics
  sumstats.test = matrix(NA, p, 3)
  
  for (i in 1:p) {
    sumstats.test[i, ] =
      c(mean(y.test[i, ]), var(y.test[i, ]), mad(y.test[i, ]))
  }
  
  ref.testing = cbind(theta1.test, theta2.test, sumstats.test)
  colnames(ref.testing) = c("theta1", "theta2", "expectation", "variance", "mad")
  
  
  # Compute the exact posterior expectations, variances and quantiles 
  # for parameters theta1 and theta2
  theta1.test.exact = rep(NA, p)
  theta2.test.exact = rep(NA, p)
  var1.test.exact = rep(NA, p)
  var2.test.exact = rep(NA, p)
  quant.theta1.test.freq = matrix(NA, p, 2)
  quant.theta2.test.freq = matrix(NA, p, 2)
  
  for (i in 1:p) {
    theta1.test.exact[i] = sum(y.test[i, ]) / (n + 1)
    var1.test.exact[i] =
      (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  ) /
      ( (n + 1) * (alpha - 1 + n / 2) )
    theta2.test.exact[i] =
      (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  ) / (alpha - 1 + n / 2)
    var2.test.exact[i] =
      (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  ) ^ 2 / ((alpha - 1 + n / 2) ^ 2 * (alpha - 2 + n / 2))
    quant.theta1.test.freq[i, ] =
      c(qnst(0.025, n + 2 * alpha, sum(y.test[i, ]) / (n + 1), sqrt(2 * (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  ) / ((n + 1) * (n + 2 * alpha) ))),
        qnst(0.975, n + 2 * alpha, sum(y.test[i, ]) / (n + 1), sqrt(2 * (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  ) / ((n + 1) * (n + 2 * alpha) ))))
    quant.theta2.test.freq[i, ] = 
      c(1 / qgamma(0.975, shape = (n + 2 * alpha) / 2, rate = (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  ) ),
        1 / qgamma(0.025, shape = (n + 2 * alpha) / 2, rate = (beta + sum((y.test[i, ] - mean(y.test[i, ])) ^ 2)/2 + n*(mean(y.test[i, ]))^2 / (2*(n+1))  )) )
  }
  
  test.exact = data.frame(mean.theta1 = theta1.test.exact,
                          var.theta1 = var1.test.exact,
                          lower.theta1 = quant.theta1.test.freq[,1],
                          upper.theta1 = quant.theta1.test.freq[,2],
                          mean.theta2 = theta2.test.exact,
                          var.theta2 = var2.test.exact,
                          lower.theta2 = quant.theta2.test.freq[,1],
                          upper.theta2 = quant.theta2.test.freq[,2])
  
  # Add noise to summary statistics simulated according 
  # to a uniform(0,1) distribution
  nNoise = 50 # or 500
  
  set.seed(3)  # for reproducibility 
  
  sumstats.noise = matrix(runif((N+p) * nNoise), N+p, nNoise) 
  ref.training = cbind(ref.training, sumstats.noise[1:N, ])
  ref.testing = cbind(ref.testing, sumstats.noise[(N+1):(N+p), ])
  
  colnames(ref.training) =
    c("theta1", "theta2", "expectation", "variance", "mad", c(1:nNoise))
  colnames(ref.testing) =
    c("theta1", "theta2", "expectation", "variance", "mad", c(1:nNoise))
  
  # Add some others summary statistics 
  y = ref.training[, 1:2]
  x = ref.training[, -c(1:2)]
  
  x =cbind(x, x[, 1] + x[, 2], x[, 1] + x[, 3], x[, 2] + x[, 3], x[, 1] + x[, 2] +
             x[, 3], x[, 1] * x[, 2], x[, 1] * x[, 3], x[, 2] * x[, 3], x[, 1] * x[, 2] *
             x[, 3])
  colnames(x) =
    c("expectation",
      "variance",
      "mad",
      c(1:nNoise),
      "sum_esp_var",
      "sum_esp_mad" ,
      "sum_var_mad",
      "sum_esp_var_mad",
      "prod_esp_var",
      "prod_esp_mad",
      "prod_var_mad" ,
      "prod_esp_var_mad")
  
  ytest = ref.testing[, 1:2]
  xtest = ref.testing[, -c(1:2)]
  
  xtest =
    cbind(
      xtest,
      xtest[, 1] + xtest[, 2],
      xtest[, 1] + xtest[, 3],
      xtest[, 2] + xtest[, 3],
      xtest[, 1] + xtest[, 2] + xtest[, 3],
      xtest[, 1] * xtest[, 2],
      xtest[, 1] * xtest[, 3],
      xtest[, 2] * xtest[, 3],
      xtest[, 1] * xtest[, 2] * xtest[, 3]
    )
  colnames(xtest) =
    c("expectation",
      "variance",
      "mad",
      c(1:nNoise),
      "sum_esp_var",
      "sum_esp_mad" ,
      "sum_var_mad",
      "sum_esp_var_mad",
      "prod_esp_var",
      "prod_esp_mad",
      "prod_var_mad" ,
      "prod_esp_var_mad")
  
  data.theta1 = data.frame(theta1 = y[,1], x)
  data.theta2 = data.frame(theta2 = y[,2], x)
  
  param.Test = data.frame(ytest)
  
  colnames(param.Test) = c("theta1", "theta2")
  
  stats.Test = data.frame(xtest)
  
  colnames(stats.Test) = colnames(x)
  colnames(param.Test) = colnames(y)
  
  return(list(x.train = x,
              y.train = y,
              x.test = stats.Test,
              y.test = param.Test,
              y.exact = test.exact))
}


dataset = homoscedastic_normal(alpha, beta, n, N, p, offset)

saveRDS(dataset, "../inst/extdata/normal_toy_model.Rds")
```



```{r, echo = F}
dataset = readRDS("../inst/extdata/normal_toy_model.Rds")
```


```{r, echo = T}
# For training
sumstats.train = as.data.frame(dataset$x.train)
theta.train = as.data.frame(dataset$y.train)

# Testing
sumstats.test = dataset$x.test
theta.test = dataset$y.test

# The exact value to find
theta.exact = dataset$y.exact
```


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
ggplot(theta.train, aes(x = theta1, y = theta2)) +
  geom_point() +
  geom_point(data = theta.test, aes(x = theta1, y = theta2), color = "red", alpha = 0.5)

```



## Training a Deep Ensemble model with adversarial training


```{r}
deepensemble_highdim = abcnn$new(theta.train,
            sumstats.train,
            sumstats.test[1:1000,],
            method = 'deep ensemble',
            scale_input = "minmax",
            scale_target = "minmax",
            num_hidden_layers = 3,
            num_hidden_dim = 256,
            epochs = 20,
            batch_size = 64,
            l2_weight_decay = 1e-4,
            epsilon_adversarial = 0.001,
            seed = 42)


deepensemble_highdim$summary()
```



```{r, eval = F}
deepensemble_highdim$fit()
```


With adversarial training, the training and validation losses are different (adversarial perturbation is applied on training only), yet the training curve is smoother after a strong perturbation at the beginning:

```{r, echo = F, eval=T}
deepensemble_highdim = load_abcnn(prefix = "../inst/extdata/deepensemble_highdim")
```


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
deepensemble_highdim$plot_training()
```



## Model performance


Evaluate performance with cross-validation (here we re-use the previous observed test set for convenience):


```{r, echo = T}
true.theta = data.frame(theta1 = theta.exact$mean.theta1,
                        theta2 = theta.exact$mean.theta2)

deepensemble_highdim$cross_validation(true.theta,
                                      sumstats.test)

deepensemble_highdim$plot_cross_validation()
```


Plot the predictions on the observed summary statistics:

```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="TabNet-ABC performance showing posterior quantile predictions compared to true parameter values."}
deepensemble_highdim$predict()

deepensemble_highdim$plot_prediction(uncertainty_type = "conformal")
```


```{r, echo = F, eval=F}
save_abcnn(deepensemble_highdim, prefix = "../inst/extdata/deepensemble_highdim")
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="Scatter plot comparing DeepEnsemble predictions to exact posterior means. The shaded region represents 95% credible intervals."}
df = deepensemble_highdim$predictions() %>%
  filter(parameter == "theta1")

df$true.theta = theta.exact[1:1000,"mean.theta1"]

ggplot(df, aes(x = true.theta, y = predictive_mean)) +
  geom_ribbon(aes(ymin = predictive_mean - overall_conformal_credible_interval,
                  ymax = predictive_mean + overall_conformal_credible_interval), 
              fill = "lightgrey", alpha = 0.5) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "True Posterior Mean", y = "Predicted Mean",
       title = "Deep Ensemble Predictive check") +
  theme_bw()
```




```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="Scatter plot comparing DeepEnsemble predictions to exact posterior means. The shaded region represents 95% credible intervals."}
df = deepensemble_highdim$predictions() %>%
  filter(parameter == "theta2")

df$true.theta = theta.exact[1:1000,"mean.theta2"]

ggplot(df, aes(x = true.theta, y = predictive_mean)) +
  geom_ribbon(aes(ymin = predictive_mean - overall_conformal_credible_interval,
                  ymax = predictive_mean + overall_conformal_credible_interval), 
              fill = "lightgrey", alpha = 0.5) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "True Posterior Mean", y = "Predicted Mean",
       title = "Deep Ensemble Predictive check") +
  theme_bw()
```


# Model Interpretability and Feature Importance

Understanding which summary statistics drive parameter estimates is crucial for scientific interpretation. The **explainn** class implements a suite of methods to provide interpretable feature attributions. 


## The `explainn()` class and methods for model explanation



The [https://bips-hb.github.io/innsight/](**innsight**) package backend provides twelve different explainability methods to interpret feature importance within an **explainn** object. Both local and global methods are provided, to interpret feature importance for a focal prediction (a single output/sample) or summarize importances across a whole dataset (multiple samples).


DeepLift (deep learning important features) is an algorithm introduced by Shrikumar et al. (2017). It's a local method for interpreting a single element output prediction $x$, given a reference $x'$, returning the contribution of each input feature from the difference of the output $y = f(x)$ and the reference output $y' = f(x')$ prediction.



```{r, echo = T, eval = T}
exp = explainn$new(deepensemble_highdim)

exp$run(data = sumstats.test[1:1000,],
        method = "deeplift")
```



```{r explainn_plot, echo = T, fig.height = 4, fig.width = 8, fig.cap="Feature importance visualization using DeepLIFT attribution methods, for the first sample output prediction. Colors indicate the contribution of each feature to the final prediction."}
exp$plot()
```


Even if DeepLift is a local method, multiple samples can be passed and importance scores summarizzed across all samples' predictions.



```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="Feature importance visualization using DeepLIFT attribution methods, summarized across the 1,000 smaples. Colors indicate the contribution of each feature to the final prediction."}
exp$plot_global()
```



The individual contributions of input variables to the subject prediction:


```{r, echo = T}
# Three dimension outpout:
# - First sample
# - 10 first summary statistics
# - the two output parameters
exp$get_result()[1,1:10,1:2]
```




SmoothGrad is another method, which can also be interpreted as local and global feature importance scores. SmoothGrad was introduced by D. Smilkov et al. (2017) and is an extension to the classical Vanilla Gradient method (see [SmoothGrad](https://bips-hb.github.io/innsight/reference/SmoothGrad.html) for more details).



```{r, echo = T, eval = T}
exp = explainn$new(deepensemble_highdim)

exp$run(data = sumstats.test[1:1000,],
        method = "smoothgrad")
```

```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="Feature importance visualization using DeepLIFT attribution methods. Colors indicate the contribution of each feature to the final prediction."}
exp$plot()
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="Feature importance visualization using DeepLIFT attribution methods. Colors indicate the contribution of each feature to the final prediction."}
exp$plot_global()
```


The individual contributions of input variables to individual predictions:


```{r, echo = T}
exp$get_result()[1:5,1:10,1]
```








# ABC Integration with TabNet

## TabNet Architecture for Summary Statistics

TabNet-ABC [@arik_tabnet_2020; @akesson_convolutional_2022] combines interpretable attention mechanisms with traditional ABC methods. The key innovation is using TabNet to learn informative summary statistics from raw simulation outputs, then applying standard ABC regression on these learned features.


## TabNet-ABC Implementation


Tabnet is trained to predict the mean parameter, which is re-used during inference as a new informative summary statistics. We leverage the Tabnet Attention map and feature importance to subset only the most informative summary statistics for the ABC inference step. They are then concatenated to the Tabnet predicted mean.


```{r}
tabnetabc = abcnn$new(theta.train[1:10000,],
            sumstats.train[1:10000,],
            sumstats.test[1:1000,],
            method = 'tabnet-abc',
            scale_input = "none",
            scale_target = "none",
            epochs = 30,
            batch_size = 64,
            tol = 0.1,
            abc_keep_original_sumstats = 8,
            abc_method = "loclinear",
            seed = 4567)


tabnetabc$summary()
```



```{r, eval = F}
tabnetabc$fit()
```


```{r, echo = F, eval=T}
tabnetabc = load_abcnn(prefix = "../inst/extdata/tabnetabc")
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
tabnetabc$plot_training()
```


```{r, eval = F, message=FALSE, results = FALSE, include=FALSE}
tabnetabc$predict()
```



```{r, echo = F, eval=F}
save_abcnn(tabnetabc, prefix = "../inst/extdata/tabnetabc")
```


## Model performance


```{r tabnet_performance, echo = T, fig.height = 4, fig.width = 8, fig.cap="TabNet-ABC performance showing posterior quantile predictions compared to true parameter values."}
tabnetabc$plot_prediction(uncertainty_type = "posterior quantile",
                          plot_type = "errorbar")
```



```{r tabnet_accuracy, echo = T, fig.height = 4, fig.width = 8, fig.cap="Scatter plot comparing TabNet-ABC predictions to exact posterior means. The shaded region represents 95% credible intervals."}
df = tabnetabc$predictions() %>%
  filter(parameter == "theta1")

df$true.theta = theta.exact[1:1000,"mean.theta1"]

ggplot(df, aes(x = true.theta, y = predictive_mean)) +
  geom_ribbon(aes(ymin = posterior_lower_ci, ymax = posterior_upper_ci), 
              fill = "lightgrey", alpha = 0.5) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "True Posterior Mean", y = "Predicted Mean",
       title = "TabNet-ABC Predictive check") +
  theme_bw()
```



<!-- ```{r, fig.height = 4, fig.width = 8, fig.align="center"} -->
<!-- tabnetabc$plot_prediction(uncertainty_type = "posterior quantile", plot_type = "errorbar") -->
<!-- ``` -->


<!-- ```{r, message=F, echo=TRUE, fig.height = 4, fig.width = 8, fig.align="center"} -->
<!-- tabnetabc$plot_posterior(5, uncertainty_type = "posterior quantile") -->
<!-- ``` -->


<!-- ```{r, message=F, echo=TRUE, fig.height = 4, fig.width = 8, fig.align="center"} -->
<!-- df = tabnetabc$predictions() %>% -->
<!--   filter(parameter == "theta1") -->

<!-- df$true.theta = theta.exact[1:1000,"mean.theta1"] -->

<!-- ggplot(df, aes(x = true.theta, y = predictive_mean)) + -->
<!--   geom_ribbon(aes(ymin = posterior_lower_ci, ymax = posterior_upper_ci), fill = "lightgrey", alpha = 0.5) + -->
<!--   geom_point() + -->
<!--   geom_smooth(method = "lm") + -->
<!--   theme_bw() -->
<!-- ``` -->




## Tabnet attention maps for model explanation


With the method **Tabnet-ABC**, built-in attention maps and **tabnet** methods are used instead of **innsight** methods:


```{r, echo = T, eval = T}
exp = explainn$new(tabnetabc)

exp$run(data = sumstats.test)
```


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
exp$plot()
```


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
exp$plot(type = "mask_agg")
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
exp$plot(type = "steps")
```



## Improve model performance by keeping the most important summary statistics



```{r}
tabnetabc_performance = abcnn$new(theta.train[1:10000,],
            sumstats.train[1:10000,],
            sumstats.test[1:1000,],
            method = 'tabnet-abc',
            scale_input = "none",
            scale_target = "none",
            epochs = 30,
            batch_size = 256,
            l2_weight_decay = 1e-3,
            tol = 0.1,
            abc_keep_original_sumstats = 0.3,
            abc_method = "ridge",
            seed = 4242)


tabnetabc_performance$summary()
```


```{r, echo = F, eval=T}
tabnetabc_performance = load_abcnn(prefix = "../inst/extdata/tabnetabc_performance")
```


```{r, eval = F}
tabnetabc_performance$fit()
```


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
tabnetabc_performance$plot_training()
```

```{r, eval = F}
tabnetabc_performance$cross_validation(theta.test[1:1000,],
                                       sumstats.test[1:1000,])
```


```{r, echo = T, fig.height = 4, fig.width = 8, fig.align="center"}
tabnetabc_performance$plot_cross_validation()
```


```{r, eval = F, message=FALSE, results = FALSE, include=FALSE}
tabnetabc_performance$predict()
```



```{r, echo = F, eval=F}
save_abcnn(tabnetabc_performance, prefix = "../inst/extdata/tabnetabc_performance")
```



```{r, echo = T, fig.height = 4, fig.width = 8, fig.cap="TabNet-ABC performance showing posterior quantile predictions compared to true parameter values."}
tabnetabc_performance$plot_prediction(uncertainty_type = "posterior quantile", plot_type = "errorbar")
```


```{r tabnet_accuracy_2, echo = T, fig.height = 4, fig.width = 8, fig.cap="Scatter plot comparing TabNet-ABC predictions to exact posterior means. The shaded region represents 95% credible intervals."}
df = tabnetabc_performance$predictions() %>%
  filter(parameter == "theta1")

df$true.theta = theta.exact[1:1000,"mean.theta1"]

ggplot(df, aes(x = true.theta, y = predictive_mean)) +
  geom_ribbon(aes(ymin = posterior_lower_ci, ymax = posterior_upper_ci), 
              fill = "lightgrey", alpha = 0.5) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "True Posterior Mean", y = "Predicted Mean",
       title = "TabNet-ABC Predictive check") +
  theme_bw()
```






# FAQ



*1. How does it work?*


**abcneuralnet** proposes a simple straightforward workflow:


1. Training. Define your data and model architecture with `abcnn$new()`. Call `abcnn$fit()` on your model. All is very user-friendly, `torch` and `luz` handle batching, backpropagation, and optimization automatically, as well as handling the device (cpu or CUDA gpu).



2. Validation. During training, `luz` evaluates and print the performance of the model on a separate validation dataset to monitor overfitting and generalization at each epoch. After training, check training and validation loss curves with `abcnn$plot_training()` to evaluate under/overfitting.



3. Testing. After training, evaluate the trained model on an unseen test set (a subset of the training data). The multi-output predictions of **abcneuralnet** allow you to evaluate the predictive performance by predicting on a test set (a simulated dataset with known ground truth). Use `abcnn$predict()` method to get predictions on both the test set and calculate custom metrics on the test set. 



4. Posterior predictive check is an additional important step to evaluate the model's performance in a Bayesian framework. It is similar to step 3, but instead of testing on a subset of the training data, you must re-generate a simulated dataset where the priors are the posterior estimated by the model. Call `abcnn$predict(posterior_check_data)` where `posterior_check_data` is your new simulated summary statistics generated by drawing parameters in the posterior. As in step 3, calculate custom metrics on the posterior predictive test set to evaluate the model's predictive performance.



5. Iteratively improve the model. Adjust hyperparameters or model architecture based on diagnostics, and re-train.



6. Infer parameters based on observed summary statistics. Call `abcnn$predict(observed_data)` where `observed_data` is your observed summary statistics.




At first, I advise to train a smaller/simpler model (around 3-5 hidden layers and between 100 and 500 neurons per layer max), on a reduced dataset (10,000 - 100,000 samples), to check first how it behaves with your dataset. A small training set will allow you to rapidly explore the hyperparameter space. Then, once you have found a model learning well, you can increase the training dataset and adjust accordingly the hyperparameters to scale up (usually `batch size` and `learning rates` have to be adjusted to account for overfitting when the sample size increases, see Table 1 for details on hyperparameters).


Most importantly, feel free to contact me for any questions and comments on your usage of the package (and raise **github issues**). I will do my best to document any troubleshooting encountered and to address bugs as fast as possible. Note that this package is under continuous development and improvements.



 If you are not familar with Neural Networks, I recommend the book **Hands-on machine learning with Scikit-Learn, Keras, and TensorFlow** from Aurélien Géron, or this free online book [https://skeydan.github.io/Deep-Learning-and-Scientific-Computing-with-R-torch/](Deep Learning and Scientific Computing with R torch) as a good introduction to the Torch ecosystem in R. See also the documentation at [https://torch.mlverse.org/](torch for R).
 


*2. Which method should I choose?*


Bayesian neural networks possess several specific characteristics that must be considered, as detailed below. For the Concrete Dropout and Deep Ensemble methods, model training can be more challenging than with conventional neural networks, since the model jointly estimates both the mean prediction and its associated variance. However, this variance component is also what makes these two methods particularly valuable in a Bayesian context. They allow the joint estimation of the model's epistemic uncertainty (uncertainty in the model itself) and the aleatoric uncertainty (uncertainty inherent in the data), effectively enabling the model to "know when it doesn't know". These capabilities are powerful features for assessing model performance and overall reliability, and they arguably justify the slightly lower predictive performance often observed compared to standard neural networks.

This behavior also implies that if your training dataset (simulations) lacks informativeness or if the model architecture is poorly suited to the observed data, the neural network will explicitly indicate its inability to learn with a large epistemic uncertainty. In such cases, training will fail, typically evidenced by a loss function that decreases only weakly over epochs or exhibits large stochastic fluctuations between epochs—underscoring the importance of closely monitoring training curves. Even when training appears successful, a misspecified model or uninformative simulations will generally lead to extremely high uncertainty estimates and wide credible intervals.

In contrast, the Monte Carlo Dropout and TabNet-ABC methods do not exhibit this "variance" issue and tend to behave similarly to conventional neural networks. The Monte Carlo Dropout approach is simple, not especially high-performing, and offers limited interpretability — estimating only epistemic uncertainty — but it remains useful for quickly testing hypotheses thanks to its ease of training. The TabNet-ABC method follows a similar principle to that described in Åkesson et al. (2022), with the key distinction that the CNN is replaced by a Transformer neural network, which performs better on tabular data. In this framework, a neural network estimates the mean posterior of the parameters, which are then used as summary statistics for a standard ABC procedure. Thus, the Bayesian component is implemented through the ABC algorithm (e.g., using the abc package), rather than being integrated into the neural network itself.

The TabNet architecture is highly effective for processing tabular data — such as the reference table in ABC — and includes a built-in attention mechanism that provides feature importance scores. Consequently, TabNet-ABC is an excellent alternative if Concrete Dropout and Deep Ensemble prove difficult to apply, or if one seeks to remain closely aligned with a more traditional ABC framework.



Choosing the appropriate method depends on several factors:

| Method | Strengths | Limitations | Computational Cost |
|--------|-----------|-------------|-------------------|
| Concrete Dropout | Principled uncertainty, Bayesian framework, Almost as efficient as Deep Ensemble for predictive performance and Credible Interval | Training complexity, heteroscedastic loss, Sensitive to statistical outliers and large variance in the training dataset (scaling the data is recommended), Epistemic uncertainty estimates can be noisy/too large when overfitting (prefer the overall uncertainty) | Medium |
| Monte Carlo Dropout | Simple implementation, fast inference | Limited uncertainty quantification (only epistemic), Less efficient than other methods, Larger credible intervals, Requires fine-tuning the dropout rate | Low |
| Deep Ensemble | Robust, adversarial training, The most efficient in predictive performance and Credible Interval, Smoother uncertainty estimates | High memory usage and training time increases with the number of networks, but inference is very fast, does not generate a full posterior (only point estimates of mean and variance) | Medium |
| TabNet-ABC | Interpretability, ABC integration, The Transformer architecture is designed to handle tabular data, built-in Attention map | Two-stage process, Requires an additional ABC procedure to estimate the full posterior, Only epistemic uncertainty | High |



*3. Should I transform the input data?*

Pre-processing and scaling the input data is crucial to achieve good performances with neural networks. It is even more true with Bayesian neural networks (BNN), as large differences in scales within or between variables can have more dramatic effects on the variance than the mean. If your model is not learning well or have an unexpected behavior, I advise you to think about the nature of your data and try to find an appropriate scaling function. Scaling the data is very easy, as scaling both the input and target data is directly implemented in the package, through the `scale_input` and `scale_target` arguments. You can choose the `minmax`, `robustscaler` or `normalization` scaling methods, or just `none` if you don't want scaling. Usually `minmax` and `robustscaler` perform well. For the scaling, the scale function learns its parameters on the training set only to avoid data leakage. If you choose to scale the target too (for example if your target is a mix of rates and continuous values), the predictions are transformed back to the original scale.



*4. How do I evaluate the training performance of my model?*

To evaluate the training of your model (how well it learned), the best approach is simply to look at the learning curves. I will not detail how to interpret these curves, there is a lot of literature about it and it is the same for every neural network. See for example [this post](https://rstudio-conf-2020.github.io/dl-keras-tf/notebooks/learning-curve-diagnostics.nb.html). The package offers all the necessary information to evaluate the learning curve, with losses estimated on three independent training, validation and evaluation sets.



*5. Which type of uncertainty must I consider?*


Once your model has passed the learning-curve diagnostic, you can move on to deeper validation by examining and comparing the epistemic and aleatoric uncertainties. These raw uncertainty components provide particularly useful insight into model behavior across samples, allowing you to identify regions where predictions are accurate and confident (low epistemic uncertainty) and regions where the model effectively "does not know", making predictions unreliable. In general, epistemic uncertainty is high in areas with few data points or outside the training distribution. This analysis also helps determine whether performance is mainly limited by the model (epistemic) or by the data (aleatoric), noting that epistemic uncertainty can often be reduced with more or better data, whereas aleatoric uncertainty reflects irreducible noise in the data. Typically, epistemic uncertainty should be noticeably lower than aleatoric uncertainty; if epistemic uncertainty is similar to or higher than aleatoric uncertainty, it indicates that the model has not learned the data-generating structure effectively.

For publications or reporting, conformal credible intervals are generally more appropriate than raw uncertainty values, because they are calibrated to achieve a specified coverage level (by default, a 95% interval). The interpretation of the Conformal Credible Intervals is similar to that of Bayesian credible intervals, while additionally offering formal coverage guarantees under suitable assumptions.

<!-- Leverage the multi-output feature -> predict on a simulated test set alongside the observed dataset -->
<!-- You can design your own evaluation procedure and metric to assess the performance of the model -->


<!-- Feature importance/selection -->






*7. How do I tune the hyperparameters of the model?*

Stick to simple, widely-used standard values. Train simpler models for test purpose, before running larger models in production. Beware than larger models can sometimes require adjustments for the learning hyperparameters when scaling up (learning rate, batch size), as larger sample sizes also have effects on overfitting and on the variance that is estimated by the model. A larger sample size affects the epistemic uncertainty, the uncertainty of the model, while the aleatoric uncertainty, the uncertainty of the data itself, should not be much affected.


The number of hidden layers and hidden dimensions (number of neurons) are important to efficiently capture complex non-linear relationships between input and output features, but with Bayesian Neural Network it is not necessary to use a large number of neurons and layers. Generally 3-6 layers with 100-500 neurons each is sufficient. Yet the most critical hyperparameters are those related to the training procedure, such as the learning rate and the batch size. The L2 weight decay only has a minimum impact, and the default value should be fine, except if you observe a strong overfitting.


```{r, echo=F}
hyperparam = data.frame(Hyperparameter = c("Number of hidden dimensions (i.e. neurons) in one layer",
                                           "Number of hidden layers",
                                           "Epochs",
                                           "Batch size",
                                           "Learning rate",
                                           "L2 weight decay",
                                           "Dropout (only for Monte Carlo Dropout)"),
                        Values = c("100-500 is generally a good range. You can try higher values if you have more than 200 summary statistics.",
                                   "3-6, deeper networks tend to be harder to train",
                                   "30-60 are generally enough to avoid overfitting",
                                   "32-256, depending mostly on the variance in your training set. Higher batch sizes have a smoothing effect on excess of variance during training.",
                                   "1e-3 is generally fine for 10,000-100,000 training samples. 1e-4 tend to be better for sample sizes >= 1,000,000.",
                                   "1e-5, minimal impact.",
                                   "0.1-0.5. Smaller values should produce smaller Credible Intervals, but at the risk of less precise or biased estimates. Higher values are more conservative."))

  
knitr::kable(hyperparam, caption = "Hyperparameters available in abcnn and standard range of values.")
```



<!-- 
*8. What to do if my model has difficulties to learn?*

Bayesian neural nets can sometimes be harder to train than regular neural nets
In concrete dropout, dropout is used at prediction time to approximate the Bayesian posterior, hence it is not efficiently used as regularization during training. This is something on which I am still working on. For example, you can use the early stopping callback to stop the training as soon as a good level of performance is achieved, before overfitting.
Tests with smaller training sets
Reduce model complexity
Reduces learning rate (slower learning), larger batch size, larger L2 weigths decay for regularization
Beware that a batch size too high or a learning rate too slow could prevent an efficient learning. As always with Neural Network, there is a significant part of engineering to find a good model, that require some steps of trial/error.

Improving gradient stability and convergence is a next step towards a more user-friendly application.

More specifically:

*9. Very large epistemic error but the aleatoric/overall is fine?*

Sign of overfitting, happens especially with training set with a large number of samples, counter-intuitive
Increase batch size. Smoothening effect.
Reduce L2 weight decay and learning rate?


*10. Huge loss and/or NaN output values during training?*

The model fails during the training step, due to an explosion of the loss value.
This is something that can happen with the methods concrete dropout and deep ensemble, because the variance component is co-estimated in the model. It usually tells you that the model has difficulties to learn due to high variance in the samples or too much complexity in the model, leading to very large loss values, then returning NaN/NA values at the next training step.
Reduce the learning rate, increase the batch size, reduce model complexity



*11. I used conformal prediction, by my C.I. coverage is different from the one expected*

Calibration with conformal prediction guarantees a coverage at least equal or superior to the expected. For many reasons the coverage can be higher than expected.


More samples in conformal prediction

The performance of the model is limited, hence a more conservative CI





*12. How much time, memory and CPU/GPU do I need?*


Torch can leverage CUDA


For faster training, increase batch size, use early stopping -->




# References

<div id="refs"></div>



# Session Information

```{r session_info}
sessionInfo()
```

